Bayesian Hyper-Heuristics

### 2023-01-22 article/sections/03-heuristics.tex
✨ (article): Heuristics section done

A \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Many different techniques have been used to train \acp{FFNN}~\cite{ref:kingma:2014}. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques~\cite{ref:nel:2021}. However, other techniques have also been used to successfully train \acp{FFNN}, including \acp{MH} such as \acf{PSO}~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE}~\cite{ref:espinal:2011} and \acfp{GA}~\cite{ref:gupta:1999}. This research takes a particular interest in both gradient-based heuristics as well as \acp{MH}.

@@

Gradient-based \index{heuristic}heuristics are optimisation techniques that make use of derivates obtained from evaluating the model being optimised. There are many variants of gradient-based \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acf{GD} framework called \acf{BP}.

In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \acs{FFNN} and the desired output. When using \acs{GD} to train \acp{FFNN}, the \index{loss function}loss function is used to adjust the weights of the \acs{FFNN} in order to minimise the error~\cite{ref:engelbrecht:2007}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} states that training of \acp{FFNN} using \acs{GD}, is done by calculating the gradient of $\epsilon$ in \textit{weight-space}, and then moving the weight vector along the negative gradient.

In the context of training \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer to the input layer, updating the weights at the output and hidden layers. The algorithm that propagates the error signal backwards is known as \acf{BP}. \Acs{BP} was popularised by~\citeauthor{ref:werbos:1994}~\cite{ref:werbos:1994}.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\mathcal{E}$, with respect to the weights at each layer.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} describes the \acs{BP} process in two steps:

\begin{itemize}
	\item \textbf{Feedforward Pass}: During this phase the output values of the \acs{FFNN} is calculated for each training pattern.

	\item \textbf{Backward Propagation}: During this phase the error signal is propagated backwards from the output layer, through the hidden layer, to the input layer of the \acs{FFNN}. Weights at the output and hidden layers are then adjusted as functions of the backpropagated error signal.
\end{itemize}

% The update step for \acs{GD} by means of \acs{BP} can be formulated as is shown in Equations~\eqref{eq:heuristics:gd:update_step_part_1} to \eqref{eq:heuristics:gd:update_step_part_4}. The general weight update step is given as

% \begin{equation}
% 	\label{eq:heuristics:gd:update_step_part_1}
% 	w_{i}(t) = w_{i}(t-1) + \Delta w_{i}(t)
% \end{equation}

% where $w_{i}$ is the weight value at index $i$, $t$ is the time step, $\Delta w_{i}(t)$ is the weight update vector at index $i$ and time step $t$. The delta weight term $\Delta w_{i}(t)$, is given as

% \begin{equation}
% 	\label{eq:heuristics:gd:update_step_part_2}
% 	\Delta w_{i}(t) = -\eta\frac{\partial \epsilon}{\partial w_{i}}
% \end{equation}

% where $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each time step. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight vector $\boldsymbol{w}$, at index $i$, is given as

% \begin{equation}
% 	\label{eq:heuristics:gd:update_step_part_3}
% 	\frac{\partial \epsilon}{\partial w_{i}} = -2(t_{i,p} - o_{i,p})\frac{\partial f}{\partial net_{p}}z_{i,p}
% \end{equation}

% where $t_{i,p}$ refers to the $i$-th element of the target value for pattern $p$, $o_{i,p}$ refers to the $i$-th element of the output value for pattern $p$, $f$ refers to the \index{activation function}activation function, $net_{p}$ refers to the \index{net input signal}net input signal for pattern $p$, and $z_{i,p}$ refers to the input value at index $i$ for pattern $p$.

% Finally, assuming the use of the \index{sigmoid}sigmoid activation function, the partial derivative of the activation function, relative to the net input signal is given as

% \begin{equation}
% 	\label{eq:heuristics:gd:update_step_part_4}
% 	\frac{\partial f}{\partial net_{p}} = o_p(1 - o_{p})
% \end{equation}

% Again, assuming the use of \acs{SSE} as the loss function, the pseudo-code implementation for the generic \acs{GD} algorithm is taken from~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:gd} below.

% \begin{algorithm}[htb]
% 	\caption{The pseudo-code algorithm for the generic \acf{GD} heuristic.}
% 	\label{algo:heuristics:gd:gd}
% 	\begin{algorithmic}
% 		\State Initialise the \acs{FFNN} weights and biases, the learning rate $\eta$ and the time step/epoch $T=0$;
% 		\While{stopping conditions are not met}
% 		\State Let $\epsilon_{T} = 0$;
% 		\For{each training pattern p}
% 		\State Calculate the output of the hidden and output layers (feedforward);
% 		\State Compute the error signals of the hidden and output layers;
% 		\State Adjust the weights of the output and hidden layers (backpropagate);
% 		\State $\epsilon_{T}$ += $[\epsilon_{T_{p}} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$;
% 		\EndFor
% 		\State $T += 1$;
% 		\EndWhile
% 	\end{algorithmic}
% \end{algorithm}

A simplified notation is proposed from which the gradient-based heuristics can be compared. Consider a simplification of the update-step for \acs{SGD} as presented in Equations~\eqref{eq:heuristics:gd:update_step_part_1} and~\eqref{eq:heuristics:gd:update_step_part_2} where $\frac{\partial \epsilon}{\partial w_{i}}$ is simply represented as the gradient term $\boldsymbol{g}$, resulting in a simplified weight update step and is given as

@@

where $\boldsymbol{w}$ refers to the weight vector, $\eta$ refers to the learning rate as before and $\boldsymbol{g}$ refers to the gradient of the error function relative to weight vector. Note that some subscripts related to layers, indices, time steps and training patterns are omitted for convenience.

In the context of this dissertation, the implementation of \acs{SGD} refers to the mini-batch training implementation of \acs{GD}. Although mini-batch training does provide a compromise between stochastic and batch training, there are still a number of challenges faced by mini-batch training. These include:

\begin{itemize}
	\item The appropriate value to use for the learning rate $\eta$, is difficult to determine and is often problem-specific. A learning rate that is too small causes premature exploitation, leading to slow and bad convergence. On the contrary, a learning rate that is too high may lead to bad learning outcomes as the heuristic keeps overshooting good minima.

	\item Learning rate schedules~\cite{ref:robbins:1951} can be introduced to dynamically change the learning rate throughout the training process, however, these schedules and their parameters have to be defined a priori and are often problem specific~\cite{ref:darken:1992}.

	\item The learning rate that has been introduced so far is applied to all elements of the weight update vector. If the training data is sparse and the features have varying frequencies, an equal update to all weight elements is inefficient. Larger weight updates are required for less frequently occurring features. Inefficient updates of all elements in the weight vectors contribute to the credit assignment problem~\cite{ref:rumelhart:1986}, common to \acs{GD} variants.

	\item It is difficult to avoid getting trapped in local minima, especially for highly non-convex loss functions used for training \acp{FFNN}. \citeauthor{ref:dauphin:2014}~\cite{ref:dauphin:2014} mentions the role of saddle points in getting trapped in local minima. Saddle points are points at which one dimensions slopes upwards, while another dimension slopes downwards. \citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that these saddle points are usually surrounded by plateaus of the same error, leading to gradients that are close to zero in all dimensions.
\end{itemize}

Alternative variants have been proposed that lead to better control over the convergence characteristics caused by \acs{GD}. The first of these \acs{GD} variants include \acs{Momentum}.

\subsubsection{Momentum}\label{sec:heuristics:gd:momentum}

Research shows that \acs{SGD} has difficulty navigating ravines~\cite{ref:sutton:1986}. Ravines are areas where the surface curves much more steeply in one dimension than in another. These ravines are common around local minima. As such, \acs{SGD} is shown to oscillate across the slopes of the ravine while only making minor progress towards the local minima.

\Acs{Momentum} ~\cite{ref:qian:1999} is a variant of \acs{SGD} that helps accelerate \acs{SGD} in the relevant direction, dampening oscillations.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that this is done by adding a fraction $\alpha$, of the weight update vector of the previous time step to the current weight update vector. An illustration of \acs{SGD} with and without momentum is given in Figure~\ref{fig:heuristics:gd:sgd_with_and_without_momentum}.

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.99\textwidth]{images/sgd_with_and_without_momentum.pdf}
	\caption{An illustration of \acf{SGD} with and without momentum taken from~\cite{ref:du:2019}.}
	\label{fig:heuristics:gd:sgd_with_and_without_momentum}
\end{figure}

The accumulation of momentum is given as

\begin{equation}
	\boldsymbol{v} = \alpha \boldsymbol{v} - \eta \boldsymbol{g}
	\label{eq:heuristics:gd:momentum_part_1}
\end{equation}

while the update step is then amended as

\begin{equation}
	\boldsymbol{w} = \boldsymbol{w} + \boldsymbol{v}
	\label{eq:heuristics:gd:momentum_part_2}
\end{equation}

By redefining the \acs{SGD} update steps as shown above in Equations~\eqref{eq:heuristics:gd:momentum_part_1} and~\eqref{eq:heuristics:gd:momentum_part_2}, the \acs{Momentum} heuristic allows for the increase of momentum for dimensions whose gradients point in the same direction, while simultaneously reducing momentum for dimensions whose gradients change direction, leading to faster convergence and less oscillation. The momentum term $\alpha$ is usually set to 0.9~\cite{ref:engelbrecht:2007, ref:ruder:2016}.

\subsubsection{Nesterov Accelerated Gradients}\label{sec:heuristics:nag}

\Acf{NAG} is a variant of the \acs{Momentum} heuristic developed by~\citeauthor{ref:sutskever:2013}~\cite{ref:sutskever:2013} and is inspired by~\citeauthor{ref:nesterov:1983}'s~\cite{ref:nesterov:1983} work on optimising convex functions. \acs{NAG} provides an improvement to the momentum accumulation term by providing a look-ahead term that better refines the weight update step.

In the \acs{NAG} heuristic, the gradient is not calculated w.r.t the current weights, but rather w.r.t the approximate future positions of the weights~\cite{ref:sutskever:2013, ref:ruder:2016}. An illustration of the weight update vector using \acs{NAG} is taken from Geoffrey Hinton's lecture on mini-batch \acs{GD}~\cite{ref:hinton:2012} and is presented in Figure~\ref{fig:heuristics:gd:nag} below.

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.95\textwidth]{images/nag.pdf}
	\caption{An illustration of the weight update vector for \acf{NAG} taken from~\cite{ref:hinton:2012}.}
	\label{fig:heuristics:gd:nag}
\end{figure}

The difference in the update step for \acs{Momentum} and the update step for \acs{NAG} as developed by~\citeauthor{ref:sutskever:2013}~\cite{ref:sutskever:2013} is described by~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} as follows. \acs{Momentum} first calculates the current gradient, represented by the small \textcolor{blue}{blue} vector in Figure~\ref{fig:heuristics:gd:nag} and then takes a large step in the direction of the updated accumulated gradient, presented by the big \textcolor{blue}{blue} vector. In contrast, \acs{NAG} first takes a big step in the direction of the previous accumulated gradient presented by the \textcolor{brown}{brown} vector. At this point the gradient is measured and \acs{NAG} makes a correction represented by the \textcolor{red}{red} vector. The complete update step is represented by the green vector. By anticipating approximate future positions of the weights, the weight update step as defined by \acs{NAG} controls the optimisation process from going too fast and results in increased responsiveness~\cite{ref:bengio:2013}. Along with the same velocity update step for \acs{Momentum}, as presented in Equation~\eqref{eq:heuristics:gd:momentum_part_1}, the \acs{NAG} weight update step is given as

\begin{equation}
	\label{eq:heuristics:gd:nag_part_2}
	\begin{split}
		\boldsymbol{w} = \boldsymbol{w} + \alpha \boldsymbol{v} - \eta \boldsymbol{g}
	\end{split}
\end{equation}

\subsubsection{Adaptive Gradients}\label{sec:heuristics:adagrad}

\Acf{Adagrad} is an adaptation of \acs{SGD} that implements a learning rate for every element in the weight vector and is developed by \citeauthor{ref:duchi:2011}~\cite{ref:duchi:2011}.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that \acs{Adagrad} adapts the learning rate to the elements of the weight vector, by  performing small updates (i.e. low learning rates) for weights associated with frequently occurring features and larger updates (i.e. high learning rates) for weights associated with infrequent features. For this reason, \acs{Adagrad} is well suited for dealing with situations where training data is sparse.

In the \acs{GD} variants presented thus far, the weight updates have been applied by making use of the same learning rate $\eta$, for all elements of the weight update vector. \Acs{Adagrad} uses a different learning rate for every weight element $w_{i}$, at every time step $t$. The weight update step for \acs{Adagrad} is given as

\begin{equation}
	\label{eq:heuristics:gd:adagrad_part_1}
	\begin{split}
		w_{t+1,i} = w_{t,i} - \frac{\eta}{\sqrt{G_{t,ii} + \epsilon}}.g_{t,i}
	\end{split}
\end{equation}

where $w_{i}$ is the $i$-th element of the weight vector, $G_{t,ii} \in \mathbb{R}^{d \times d}$ is a diagonal matrix where each diagonal element $i,i$, is the sum of the squared gradients w.r.t. $w_{i}$, up to time step $t$ and $\epsilon$ is a smoothing term that avoids division by zero and is set to an insignificant value such as $1 \times 10^{-8}$. Since $\boldsymbol{G_{t}}$ contains the sum of the squares of the gradients along its diagonal, the weight update step can be vectorised and updated using the matrix-vector product, represented by $\odot$, between $G_{t,i}$ and $g_{t,i}$. The simplified update step as implemented by \acs{Adagrad} is given as

\begin{equation}
	\label{eq:heuristics:gd:adagrad_part_2}
	\begin{split}
		w_{t+1,i} = w_{t,i} - \frac{\eta}{\sqrt{G_{t,i} + \epsilon}} \odot g_{t,i}
	\end{split}
\end{equation}

\Ac{Adagrad}'s main benefit is that it does not require manual tuning of the learning rate $\eta$. However, a problem with \acs{Adagrad} is in the accumulation of squared gradients in the denominator, represented by $\boldsymbol{G_{t}}$. Since every term that is added is positive, $\boldsymbol{G_{t}}$ keeps growing, leading to a situation where the learning rate shrinks to the point that the heuristic is no longer able to learn.

\subsubsection{Adaptive Learning Rate}\label{sec:heuristics:adadelta}

\citeauthor{ref:zeiler:2012}~\cite{ref:zeiler:2012} presents \acs{Adadelta}, an improvement of \acs{Adagrad} that eliminates the accumulation of squared gradients in the denominator.~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that instead of accumulating all past squared gradients as in the case of \acs{Adagrad}, \acs{Adadelta} restricts the window of accumulation to a window with a fixed size $W$. However, storing $W$ previous squared gradients is very inefficient. Instead, the accumulation of gradients over $W$ steps is defined recursively as a decaying average of all past squared gradients. The moving average of squared gradients at time step $t$, denoted by $E[\boldsymbol{g}^{2}]_{t}$  depends on a fraction $\alpha$ of the previous average of squared gradients and the current gradient, similar to the update step for \acs{Momentum}. Also similar to \acs{Momentum}, $\alpha$ is set to 0.9. The update step for $E[\boldsymbol{g}^{2}]_{t}$ is given as

\begin{equation}
	\label{eq:heuristics:gd:adadelta_part_1}
	\begin{split}
		E[\boldsymbol{g}^{2}]_{t} = \alpha E[\boldsymbol{g}^{2}]_{t - 1} + (1 - \alpha)\boldsymbol{g}_{t}^{2}
	\end{split}
\end{equation}

To illustrate the difference between \acs{Adagrad} and \acs{Adadelta}, the \acs{SGD} weight update step for \acs{Adagrad}, as presented in~\eqref{eq:heuristics:gd:adagrad_part_2} is rewritten such that the diagonal matrix $\boldsymbol{G_{t}}$, is replaced with the moving average of squared gradients and is given as

\begin{equation}
	\label{eq:heuristics:gd:adadelta_part_2}
	\begin{split}
		\boldsymbol{w}_{t+1} = \boldsymbol{w}_{t} - \frac{\eta}{\sqrt{E[\boldsymbol{g}^{2}]_{t} + \epsilon}} \boldsymbol{g}_{t}
	\end{split}
\end{equation}

Since the denominator is the \acf{RMS} error criterion of the gradient, the criteria short-hand notation is used and is given as

\begin{equation}
	\label{eq:heuristics:gd:adadelta_part_3}
	\begin{split}
		\boldsymbol{w}_{t+1} = \boldsymbol{w}_{t} - \frac{\eta}{RMS[\boldsymbol{g}]_{t}} \boldsymbol{g}_{t}
	\end{split}
\end{equation}

~\citeauthor{ref:zeiler:2012}~\cite{ref:zeiler:2012} noted that the units of the update step as presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} do not match. The weight update vector should have the same hypothetical units as the weight vector. To fix the aforementioned problem, another exponentially decaying average is defined in terms of squared weight updates and is given as

\begin{equation}
	\label{eq:heuristics:gd:adadelta_part_4}
	\begin{split}
		E[\Delta \boldsymbol{w}^{2}]_{t} = \alpha E[\Delta \boldsymbol{w}^{2}]_{t - 1} + (1 - \alpha)\Delta \boldsymbol{w}_{t}^{2}
	\end{split}
\end{equation}

The \acf{RMS} error of weight updates is given as

\begin{equation}
	\label{eq:heuristics:gd:adadelta_part_5}
	\begin{split}
		RMS[\Delta \boldsymbol{w}]_{t} = \sqrt{E[\Delta \boldsymbol{w}^{2}]_{t} + \epsilon}
	\end{split}
\end{equation}

Note that $RMS[\Delta \boldsymbol{w}]_{t}$ is unknown at time step $t$. $RMS[\Delta \boldsymbol{w}]_{t}$ is approximated with the \acf{RMS} of weight updates until the previous time step. The learning rate in Equation~\eqref{eq:heuristics:gd:adadelta_part_3} is then replaced. The weight update step for \acs{Adadelta} is then concluded and is given as

\begin{equation}
	\label{eq:heuristics:gd:adadelta_part_6}
	\begin{split}
		\boldsymbol{w}_{t+1} = \boldsymbol{w}_{t} - \frac{RMS[\Delta \boldsymbol{w}]_{t-1}}{RMS[\boldsymbol{g}]_{t}} \boldsymbol{g}_{t}
	\end{split}
\end{equation}

The main advantage of \acs{Adadelta} is in removing the learning rate, which eliminates the need for hyper-parameter tuning of the learning rate.

\subsubsection{Root Mean Squared Error Propagation}\label{sec:heuristics:rmsprop}

\Acf{RMSProp}, presented by~\citeauthor{ref:hinton:2012}~\cite{ref:hinton:2012} is similar to \acs{Adadelta}  and was developed independently around the same time. \acs{RMSProp} is the same as the first weight update vector for \acs{Adadelta}, presented in Equation~\eqref{eq:heuristics:gd:adadelta_part_2}. A disadvantage of \acs{RMSProp} is that it still includes the learning rate term $\eta$ that needs to be tuned. For \acs{RMSProp},~\citeauthor{ref:hinton:2012}~\cite{ref:hinton:2012} suggests $\alpha$ be set to 0.9 and $\eta$ be set to $0.001$.

\subsubsection{Adaptive Moments Estimation}\label{sec:heuristics:adam}

\Acf{Adam} is another variant of \acs{SGD} that includes adaptive learning rates and is presented by~\citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014}. In addition to storing an exponentially decaying average of past squared gradients like \acs{Adadelta} and \acs{RMSProp}, \acs{Adam} also stores an exponentially decaying average of past gradients, similar to \acs{Momentum}~\cite{ref:ruder:2016}.~\citeauthor{ref:heusel:2017}~\cite{ref:heusel:2017} uses the analogy that, if \acs{Momentum} can be seen as a ball running down a slope, then \acs{Adam} behaves like a heavy ball with friction, which prefers flat minima in the error surface. The decaying averages for past gradients and past squared gradients is given in Equations~\eqref{eq:heuristics:gd:adam_part_1} and~\eqref{eq:heuristics:gd:adam_part_2} respectively.

\begin{equation}
	\label{eq:heuristics:gd:adam_part_1}
	\begin{split}
		\boldsymbol{m}_{t+1} = \beta_{1}\boldsymbol{m}_{t} + (1 - \beta_{1})\boldsymbol{g}_{t}
	\end{split}
\end{equation}

\begin{equation}
	\label{eq:heuristics:gd:adam_part_2}
	\begin{split}
		\boldsymbol{v}_{t+1} = \beta_{2}\boldsymbol{v}_{t} + (1 - \beta_{2})\boldsymbol{g}^{2}_{t}
	\end{split}
\end{equation}

In Equations~\eqref{eq:heuristics:gd:adam_part_1} and~\eqref{eq:heuristics:gd:adam_part_2} above, $\beta_{1}$ and $\beta_{2}$ are decay rates, similar to $\alpha$ for \acs{Momentum}. \citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014} suggest default values $\beta_{1}=0.9$, $\beta_{2}=0.999$ and $\epsilon = 1 \times 10^{-8}$.

\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} mentions that $\boldsymbol{m}_{t}$ and $\boldsymbol{v}_{t}$ presented above are estimates of the first moment (the mean) and the second moment (the uncentered variance) of the gradients respectively. ~\citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014} mentions that because $\boldsymbol{m}_{t}$ and $\boldsymbol{v}_{t}$ are initialised to be vectors of 0's, they are biased towards 0. The aforementioned bias is especially prominent during the initial time steps and/or when the decay rates $\beta_{1}$ and $\beta_{2}$ are small ($\beta_{1}$ and $\beta_{2}$ are close to 1). The bias-corrected first and second moment estimates are presented in Equations~\eqref{eq:heuristics:gd:adam_bc_first_moment} and~\eqref{eq:heuristics:gd:adam_bc_second_moment} respectively.

\begin{equation}
	\label{eq:heuristics:gd:adam_bc_first_moment}
	\begin{split}
		\hat{\boldsymbol{m}}_{t} = \frac{\boldsymbol{m}_{t}}{1 - \beta^{t}_{1}}
	\end{split}
\end{equation}

\begin{equation}
	\label{eq:heuristics:gd:adam_bc_second_moment}
	\begin{split}
		\hat{\boldsymbol{v}}_{t} = \frac{\boldsymbol{v}_{t}}{1 - \beta^{t}_{2}}
	\end{split}
\end{equation}

The \acs{Adam} weight update rule is then given as

\begin{equation}
	\label{eq:heuristics:gd:adam}
	\begin{split}
		\boldsymbol{w}_{t+1} = \boldsymbol{w}_{t} - \frac{\eta}{\sqrt{\hat{\boldsymbol{v}}_{t}} + \epsilon}\hat{\boldsymbol{m}}_{t}
	\end{split}
\end{equation}

@@

Gradient-based heuristics are sensitive to the problem that they are applied to, with hyper-parameter selection often dominating the research focus~\cite{ref:bengio:2000, ref:feurer:2019}.~\citeauthor{ref:blum:2003}~\cite{ref:blum:2003} mention that since the 1980's, a new kind of approximate algorithm has emerged which tries to combine basic heuristic methods in higher level frameworks aimed at efficiently and effectively exploring a search space. These methods are referred to as \index{meta-heuristic}\textit{meta-heuristics}.

This section aims to introduce the concept of \acp{MH}, with focus on population-based meta-heuristics. Three well known \acp{MH}, that have been shown to train \acs{FFNN}, are presented. These \acp{MH} include \acf{PSO}, \acf{DE} and \acfp{GA}.~\citeauthor{ref:carvalho:2006}~\cite{ref:carvalho:2006} compared various \acs{PSO} variants for training \acp{FFNN}.~\citeauthor{ref:espinal:2011}~\cite{ref:espinal:2011} compared \acs{DE} and \acs{PSO} when applied to \acs{FFNN} training and~\citeauthor{ref:gupta:1999}~\cite{ref:gupta:1999} compared \acs{BP} to a \acs{GA} for training \acp{FFNN}.

The term \acf{MH} was first introduced by~\citeauthor{ref:glover:1986}~\cite{ref:glover:1986} in 1986 and is derived from the composition of two Greek words. \index{heuristic}Heuristic derives from the verb \textit{heuriskein} which means ``to find''. The prefix, \textit{meta}, means ``beyond, in an upper level''. \Acp{MH} were often called \textit{modern heuristics}~\cite{ref:reeves:1993}.~\citeauthor{ref:blum:2003}~\cite{ref:blum:2003} mention that there is a debate as to what the formal definition of \acp{MH} is and suggests the definition of \acp{MH} to be ``high level strategies for exploring search spaces by using different methods''.

The biggest difference between \acp{MH} and gradient-bases heuristics is that
\acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN} during training and is not limited to information about the search space~\cite{ref:blum:2003}.

\subsubsection{Particle Swarm Optimisation}\label{sec:heuristics:mh:pso}

\Acf{PSO} is a stochastic population-based search algorithm based on the social behaviour of birds in a flock~\cite{ref:kennedy:1995}. By definition, the \acs{PSO} heuristic is nature-inspired.  \Acp{PSO} were first presented  by~\citeauthor{ref:kennedy:1995}\cite{ref:kennedy:1995}. Kennedy and Eberhart first applied \acp{PSO} to train of \acp{FFNN}~\cite{ref:eberhart:1995, ref:kennedy:1997}. The application of \acp{PSO} in the context of training \acs{FFNN} have been widely studied~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}. This section aims to provide the details of \acs{PSO} implementation.

This dissertation uses the term \textit{entity} for candidate solutions and a \textit{population} for a collection of entities, in the general context of population-based \acp{MH}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} mentions that in \acp{PSO} individual candidate solutions are referred to as \textit{particles} and the population is referred to as a \textit{swarm}. These particles are ``flown'' through a hyper-dimensional search space. Changes in particle position is due to social-psychological tendencies of individuals to emulate the success of other individuals. The changes in the particle position are then influenced by the experience or knowledge of the particle's neighbours. The social behaviour of particles is modelled such that they stochastically return to previously successful regions in the search space.

\citeauthor{ref:vanwyk:2014}~\cite{ref:vanwyk:2014} mentions that the swarm is usually arranged in a predefined structure, called a \textit{neighbourhood topology} that governs the communication between particles. Two specific configurations of neighbourhood topologies that exists are referred to as \textit{local best (lbest)} \acs{PSO} and \textit{global best (gbest)} \acs{PSO}. There are two main differences between the two approaches in terms of their convergence characteristics~\cite{ref:eberhart:1996}. These include:

\begin{itemize}
	\item Due to the larger particle interconnectivity of \textit{gbest} \acs{PSO}, the heuristic converges faster than with \textit{lbest} \acs{PSO}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} mentions that faster convergence comes at a cost of less diversity.

	\item As a consequence of larger diversity, the \textit{lbest} \acs{PSO} is less susceptible to getting trapped in local minima.
\end{itemize}

\citeauthor{ref:shi:1998}~\cite{ref:shi:1998} proposed a modification of the original \acs{PSO} as was presented by~\citeauthor{ref:kennedy:1995}~\cite{ref:kennedy:1995}. Their implementation focuses on the \textit{gbest} \acs{PSO} with \textit{inertia} weights. This dissertation focuses on the \citeauthor{ref:shi:1998}~\cite{ref:shi:1998} implementation of \textit{global best} \acs{PSO} and particles in this specific implementation has a number of properties associated with them~\cite{ref:vanwyk:2014}. These include:

\begin{itemize}
	\item \textbf{Position}: Refers to the candidate solution that is represented by the particle and defines the particle position within the optimisation problem's hyper-dimensional solution space. Let the current position vector, for particle $i$, at time step $t$ be denoted by $\boldsymbol{x}_{i}(t)$. Let $I$ denoted the population size and $J$ denote the search space dimensionality.

	\item \textbf{Velocity}: Represents a step size for the particle in the search space. The velocity at vector, for particle $i$, at time step $t$ is denoted $\boldsymbol{v}_{i}(t)$.

	\item \textbf{Solution Quality}: Refers to the evaluation of the particle's position with respect to the objective function. Let $f(\boldsymbol{x}_{i}(t))$ denote the quality of the solution represented by the particle's position.

	\item \textbf{Personal Best Position}: Refers to a cognitive memory construct, where each particle keeps track of their personal best position found during optimisation thus far. The personal best position is denoted $\boldsymbol{y}_{i}(t)$.

	\item \textbf{Global Best Position}: Refers to a social memory construct, where each particle has a reference to the best solution found in the particle's neighbourhood thus far. In the case of \text{gbest} \acs{PSO}, the global best position is the best position of the entire swarm. The global best position is denoted $\boldsymbol{\hat{y}}_{i}(t)$.
\end{itemize}

During initialisation, particles are randomly placed within the search space by sampling from a uniform distribution such that $\boldsymbol{x}_{i} \sim U(\boldsymbol{x}_{min}, \boldsymbol{x}_{max})$ and the velocity is set to 0. At time step 0, the particle's initial position is set to be the particle's personal best solution such that $\boldsymbol{y}_{i}(0) = \boldsymbol{x}_{i}(0)$. The particle's update step is then broken into two parts, including a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position}.

\begin{equation}
	\label{eq:heuristics:pso:velocity}
	\begin{split}
		v_{ij}(t+1) = wv_{ij}(t) + c_{1}r_{1_{j}}(t)[y_{ij}(t) - x_{ij}(t)] + c_{2}r_{2_{j}}(t)[\hat{y}_{ij}(t) - x_{ij}(t)]
	\end{split}
\end{equation}

\begin{equation}
	\label{eq:heuristics:pso:position}
	\begin{split}
		x_{ij}(t+1) = x_{ij}(t) + v_{ij}(t+1)
	\end{split}
\end{equation}

In Equations~\eqref{eq:heuristics:pso:velocity} and~\eqref{eq:heuristics:pso:position} above, $i$ refers to the $i$-th particle in the swarm and $j$ refers to the $j$-th dimension of  the particle's position. The velocity update step consists of:

>>>>

A \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Many different techniques have been used to train \acp{FFNN} \cite{ref:kingma:2014}. At the time of writing, the majority of work that is published around the training of \acp{FFNN} involves the use of gradient-based techniques \cite{ref:nel:2021}.

@@

Gradient-based \index{heuristic}heuristics are optimisation techniques that make use of derivates obtained from evaluating the \acs{ANN} being trained. In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \acs{ANN} and the desired output. When using \acs{GD} to train \acp{ANN}, the \index{loss function}loss function is used to adjust the weights of the \acs{ANN} in order to minimise the error \cite{ref:engelbrecht:2007}.

The simplest type of \acs{GD} algorithm is referred to as \acs{SGD}, which implements a gradient-based weight update step for each training pattern. The weight update step and is given as

@@

where $\boldsymbol{w}$ refers to the weight vector, $\eta$ refers to the learning rate, which controls the step size and $\boldsymbol{g}$ refers to the gradient of the error function relative to weight vector.

There are many variants of gradient-based \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acf{GD} framework that propagates the error signal backwards through the \acs{ANN}. This algorithm is known as \acf{BP}. \Acs{BP} was popularised by \citeauthor{ref:werbos:1994} \cite{ref:werbos:1994}. \citeauthor{ref:engelbrecht:2007} \cite{ref:engelbrecht:2007} describes the \acs{BP} process in two steps. During the forward pass step, the output values of the \acs{FFNN} is calculated for each training pattern. Then, during the backward propogation step, the error signal is propagated backwards from the output layer, through the hidden layer, to the input layer of the \acs{FFNN}. Weights at the output and hidden layers are then adjusted as functions of the backpropagated error signal.

In the context of this research, the implementation of \acs{SGD} refers to the mini-batch training implementation of \acs{GD}, where a small batch of training patterns are fed to the \acs{FFNN} at once and the error function is aggregated across all training patterns.

Alternative variants have been proposed that lead to better control over the convergence characteristics caused by \acs{SGD}. This research focusses on a number of these variants that include \Acs{Momentum} \cite{ref:qian:1999}, \Acf{NAG} \cite{ref:sutskever:2013}, \Acf{Adagrad} \cite{ref:duchi:2011}, \acs{Adadelta} \cite{ref:zeiler:2012}, \Acf{RMSProp} \cite{ref:hinton:2012} and \Acf{Adam} \cite{ref:kingma:2014}.

@@

Gradient-based heuristics are sensitive to the problem that they are applied to, with hyper-parameter selection often dominating the research focus \cite{ref:bengio:2000, ref:feurer:2019}. \citeauthor{ref:blum:2003} \cite{ref:blum:2003} mention that since the 1980's, a new kind of approximate algorithm has emerged which tries to combine basic heuristic methods in higher level frameworks aimed at efficiently and effectively exploring a search space. These methods are referred to as \index{meta-heuristic}\textit{meta-heuristics}.

The biggest difference between \acp{MH} and gradient-bases heuristics is that
\acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN} during training and is not limited to information about the search space \cite{ref:blum:2003}. Similar to gradient-based heuristics, a number of different meta-heuristics have been used to successfully train \acp{FFNN} \cite{ref:rakitianskaia:2012, ref:vanwyk:2014, ref:espinal:2011, ref:gupta:1999}. This research takes a particular interest in population-based \acp{MH} that have been used to successfully train \acp{FFNN}. These include \acp{PSO} \cite{ref:shi:1998}, \acf{DE} \cite{ref:price:2006} and \Acp{GA} \cite{ref:fraser:1957}.
